Signatures of Foreground Sky in Power Spectra of Redshifted Neutral Hydrogen from the Epoch of Reionization

### 2014-09-10 foreground_delay_spectrum_modeling.tex
edits in section 5, and conclusion. also a few more typos here and there

Foregrounds leak into the EoR H{\sc i} power spectrum by either having intrinsic spectral variability, high instrumental chromaticity, or through approximations in power spectrum analysis. 
%Here we focus on the instrumental effects while. 
To decouple from possible analysis effects, such as those pointed out by \citet{haz13}, we compute power spectra using the individual antenna pair based approach of \citet{pob13,par14}.  This approximates the power spectrum as the inverse fourier transform of the spectra generated by the correlator.

@@

We define the {\it delay spectrum} $V_\eta(\overline{\mathbf{u}},\eta)$ to be the inverse Fourier transform of $V_f(\overline{\mathbf{u}},f)$ along the frequency coordinate:
\begin{align}\label{eqn:delay-transform}
  V_\eta(\overline{\mathbf{u}},\eta) &= \int V_f(\overline{\mathbf{u}},f)\,W'_f(f)\,e^{i2\pi\eta f}\,\dif f,
\end{align}
where, $W'_f(f)$ is a frequency window weighting function which can be chosen to control the quality of the delay spectrum \citep{thy13,ved12}. $\eta$ represents the signal delay between the antenna pairs in an interferometer given by:
\begin{equation}\label{eqn:delay}
  \eta = \frac{\overline{\mathbf{x}}\cdot\overline{\mathbf{l}}}{c} = \frac{\overline{\mathbf{u}}\cdot\overline{\mathbf{l}}}{f}.
\end{equation}
The terms delay and lag are used interchangeably in this paper.

@@

In order to make a quantitiative comparison of delay spectra obtained with the MWA data and our simulations, we consider the following uncertainties. Our foreground models are derived from other higher frequency catalogs and sky maps. The inherent spread in spectral index increases the uncertainty while predicting fluxes at the observed frequency. Using simple error propagation, the fractional error in the delay spectrum caused by the spread in spectral index is $\sim \ln(f_\textrm{orig}/f)\,\Delta\alpha_\textrm{sp}$, where, $f_\textrm{orig}$ is the original frequency at which the catalog or map was created, $f=185$~MHz is the observed frequency, and $\Delta\alpha_\textrm{sp}$ is the spread (HWHM) in spectral index. From \citet{mau03}, we assume $\Delta\alpha_\textrm{sp} \approx 0.35$ for compact sources from NVSS and SUMSS catalogs. Although the model of \citet{deo08} yields a spectral index per direction on the sky, we assume similar uncertainties exist in spectral indices of our diffuse sky model as well, which is predominantly derived from 408~MHz map of \citet{has82}. Thus, fractional errors in delay spectrum from compact foreground objects and diffuse emission are $\sim$0.7 and $\sim$0.3 respectively. In addition, delay spectra from simulations and data each have uncertainties due to thermal noise with rms, $\Delta V^\textrm{N}_\eta(|\overline{\mathbf{x}}|,\eta) \sim$ 1.4$\times 10^7$~Jy~Hz, where the superscript N stands for thermal noise. We estimate the ratio of delay spectra from data and simulations as, $\rho = |V^\textrm{D}_\eta(|\overline{\mathbf{x}}|,\eta)|\,/\,|V^\textrm{S}_\eta(|\overline{\mathbf{x}}|,\eta)|$, where superscripts D and S denote data and simulation respectively. % Using the aforementioned uncertainties added in quadrature, we estimate the approximate expected uncertainty in $\log_{10}\rho$ and denote it by $\Delta\,\log_{10}\rho$.

@@

Figure~\ref{fig:data-sim-ratio} shows histograms of distribution of $\log_{10}\rho$ for the {\it off--zenith} (top panel) and {\it zenith} (bottom panel) pointings respectively inside the {\it foreground wedge}. The median absolute deviation from these distributions correspond to $\sim$~90\% fractional difference between data and modeling on average in either case. This is in line with aforementioned uncertainties in foreground models and thermal noise in measurements. Currently we are significantly limited by unavailability of all--sky foreground models at the observed frequency of 185~MHz. Availability of an accurate foreground model will improve the agreement of simulations with observed data.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/delta_array_histogram_wedge_sim_data_log_ratio_0.3m_ground_custom_gaussian_FG_model_asm_all_sky_nside_64_Tsys_90.0K_185.0_MHz_30.7_MHz_bnw2.0.eps}
\caption{Histogram of logarithm of ratio of observed to simulated visibilities restricted to the {\it foreground wedge} for the {\it off--zenith} (top) and {\it zenith} (bottom) pointings. Median absolute deviations of $\log_{10}\rho$ are 0.29 and 0.27 respectively. A value of zero indicates when the ratio $\rho=1$. \label{fig:data-sim-ratio}}
% \includegraphics[width=\linewidth]{figures/v1_0/delta_array_histogram_wedge_sim_data_snr_log_ratio_0.3m_ground_custom_gaussian_FG_model_asm_all_sky_nside_64_Tsys_90.0K_185.0_MHz_30.7_MHz_bnw2.0.eps}
% \caption{Significance of logarithm of ratio of observed to simulated visibilities relative to expected uncertainty restricted to the {\it foreground wedge} for the {\it off--zenith} (top) and {\it zenith} (bottom) pointings. Probability densities and cumulative distributions are shown in black and gray respectively. Region inside the dotted vertical lines denotes when the ratio, $\rho$, is consistent with unity subject to expected uncertainty. Also indicated in percentages are confidence levels estimated conservatively for such a consistency. They are $\sim$~90\% and $\sim$~95\% for {\it off--zenith} and {\it zenith} pointings respectively. \label{fig:data-sim-ratio-snr}}
\end{figure}

@@

Having established that results from modeling match those from data, we proceed to examine in further detail the signatures seen in the modeled delay spectra.

@@

\item {\it Sky Model}: Our model of the sky consists of bright compact sources and diffuse emission on diverse spatial scales as seen from figures~\ref{fig:DSM} and \ref{fig:CSM}. It also consists of localized regions of strong emission such as the Galactic plane and Galactic center. The resulting sky emission is anisotropic. In fact, the patch of sky for MWA observations is chosen from regions of low foreground emission. In our detailed study, we have divided the transiting sky into four different ``bow-tie'' shaped regions with equal areas around zenith as the origin. This results in the ability to attribute the features in the delay spectrum to eight different sectors of the sky.

\item {\it Antenna Pair Orientation}: Since the spatial structure of our foreground model is not expected to be isotropic, we divide our antenna pairs by their orientation ($\theta_\textrm{b}$) measured counter-clockwise from East. We use the following bins: $-67\fdg 5\le \theta_\textrm{b} < -22\fdg 5$, $-22\fdg 5\le \theta_\textrm{b} < 22\fdg 5$, $22\fdg 5\le \theta_\textrm{b} < 67\fdg 5$, and $67\fdg 5\le \theta_\textrm{b} < 112\fdg 5$. The bin centers are oriented towards South-East, East, North-East, and North respectively. Since delays depend on antenna pair orientation, binning by interferometer orientation allows us to match delay spectrum features to different sky directions.

\item {\it Tile Pointing and Power Pattern}: Since the MWA tiles are steered electronically, the power pattern of the tile changes in any observing mode that tracks the source. Our observations consist of a combination of allowing the sky to drift and tracking the sky. In our study, we take into account the effect of changing power pattern of the antennas on the delay spectrum (figure~\ref{fig:power_pattern}). The LST of our data ranges from $\sim$~21 hours through $\sim$~2 hours, and for our analysis we chose {\it off--zenith} and {\it zenith} pointings shown in figure~\ref{fig:pointings}.

\item {\it Instrumental Passband}: The instrumental passband characteristics have a direct effect on the delay transform owing to a Fourier relationship between the two. Due to excision of noisy and aliased edge channels repeated in each coarse channel of our passband, we saw repeated patterns of the {\it foreground wedge} in the delay spectra. As already mentioned earlier, we have used the deconvolution algorithm ({\it CLEAN}) to rid the delay spectra of band shape effects. Since this paper focuses on studying the effects of foreground on the delay spectra, we do not pursue the characterization of artifacts from the band shape and imperfections in {\it CLEAN} deconvolution.

@@

The foreground signatures we henceforth identify in the delay spectrum are characterized as arising out of the aforementioned factors. Due to numerous combinations of such factors, we use unique annotations to characterize these signatures.

It may be noted that numerous features overlap at varying levels of significance as a result of various combinations of parameters. We assign the features to their predominant causes. Secondly, we have used noiseless cases to clearly illustrate the observed foreground signatures. With the addition of noise in the visibilities which is subject to observing time, some of the weaker features may not be as prominently visible. Since the foreground signatures are far too numerous and they reside in a vast volume of parameter space such as antenna pair orientation, power pattern, patch of sky under observation, and instrumental configuration, we highlight only some examples of the most notable features observed in the delay spectrum of foreground emission.

\subsection{Diffuse Foregrounds}\label{sec:diffuse}

@@

\subsubsection{Diffuse Emission}\label{sec:diffuse-features}

Diffuse emission from the main lobe of the power patterns manifests as a branch at the top with positive delays and a zero--delay horizontal branch in the {\it off--zenith} and {\it zenith} pointings respectively, seen in figure~\ref{fig:noiseless-dsm-delay-spectrum}. The former is seen at positive delays because the main lobe of the power pattern is centered eastward of zenith, whereas in the latter it is centered at zenith. The horizontal arm in the {\it zenith} pointing fades away on antenna spacings wider than $\gtrsim 125$~m indicating that the diffuse emission is devoid of spatial structures on scales $\lesssim$~0\fdg 75.

@@

Figure~\ref{fig:noiseless-csm-delay-spectrum} shows the delay spectrum arising out of our compact objects sky model for {\it off--zenith} and {\it zenith} pointings (shown in figure~\ref{fig:CSM}) in the top and bottom panels respectively.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/delta_array_multi_baseline_CLEAN_noiseless_visibilities_0.3m_ground_custom_gaussian_FG_model_csm_all_sky_nside_64_Tsys_90.0K_185.0_MHz_30.7_MHz_bnw2.0.eps}
\caption{Delay spectrum amplitude, $|V_\eta(|\overline{\mathbf{x}}|,\eta)|$ (in units of Jy~Hz), obtained with simulations for {\it off--zenith} (top) and {\it zenith} (bottom) pointings for the compact objects foreground model without any thermal noise. White lines mark the boundaries of {\it foreground wedge} determined by the horizon delay limit and antenna spacing. The logarithmic color scale is identical to that in figure~\ref{fig:noiseless-dsm-delay-spectrum}. Emission from compact objects gives rise to centrally concentrated features in the {\it foreground wedge}.\label{fig:noiseless-csm-delay-spectrum}}
\end{figure}

@@

\subsubsection{Compact Emission in Power Pattern Main Lobe}\label{sec:csm-main-lobe}

Since the main lobe of the power pattern in the {\it off--zenith} pointing is centered eastward of zenith, bulk of the compact foreground emission is seen in a branch with positive delay but short of the positive horizon delay limit. Compact emission from the same patch of sky is seen as a bright horizontal arm at zero delay in the {\it zenith} pointing since the main lobe of the power pattern is centered at zenith.

@@

\subsection{All--Sky Composite Foreground Model}\label{sec:composite}

@@

Excluding the bright emission from the Galactic center which is heavily dependent on the sky location pointed to by the antenna, delay spectrum resulting from the all--sky composite foreground model consists of signatures already noted in the diffuse and compact foreground cases. This is illustrated by a microscopic examination of the {\it zenith} pointing in our study.

Figure~\ref{fig:pitchfork-baselines} shows delay spectra of three antenna pairs of different antenna spacings oriented northward during the {\it zenith} pointing. The diffuse, compact, and composite components are are shown as dotted, dashed, and solid lines respectively. The horizon delay limits are shown as vertical dot--dashed lines. The gray shaded area denotes the envelope of expected uncertainty in the delay spectrum. This uncertainty is dominated by the spread in spectral index of compact foreground objects inside the {\it foreground} wedge (between the horizon delay limits) proprotional to the delay spectrum from compact objects, and by thermal noise outside the horizon delay limits.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/delta_array_3_baseline_comparison_visibilities_0.3m_ground_custom_gaussian_FG_model_csm_all_sky_nside_64_Tsys_90.0K_185.0_MHz_30.7_MHz_bnw2.0.eps}
\caption{Absolute value of delay spectrum, $|V_\eta(\overline{\mathbf{x}},\eta)|$, for three chosen antenna spacings, $\sim$84~m ({\it top}), $\sim$105~m ({\it middle}), and $\sim$170~m ({\it bottom}). The antenna spacing vector is specified in each panel along East, North and Upward directions. The dotted, dashed and solid lines denote contributions from diffuse emission, compact objects and composite foreground model respectively. The dot--dashed lines mark the horizon delay limits. The shaded region denotes the envelope of uncertainty around expected values for composite model. This uncertainty is dictated by uncertainties in spectral indices of foreground models inside the delay horizon delay limits. Outside the horizon delay limits, it is predominantly due to thermal noise. Compact objects dominate the central regions of power spectrum while both components, especially diffuse emission on short antenna spacings, dominate near the horizon delay limits, giving rise to a {\it pitchfork}--shaped structure. \label{fig:pitchfork-baselines}}
\end{figure}

@@

Near the horizon delay limits at shorter antenna spacing, diffuse emission is brighter than that from compact objects. With increasing antenna spacing, although diffuse and compact components of emission become comparable, diffuse emission does not decrease as rapidly as seen at zero--delay. In fact, even with widely spaced antennas, diffuse emission in the delay spectrum near the horizon delay limits exceeds that in the main lobe by more than two orders of magnitude. This is due to apparent shortening of antenna spacing owing to projection effects while receiving emission from wide off--axis directions. This demonstrates the edge--heavy {\it two--pronged fork} feature discussed in \S\ref{sec:diffuse-long-baselines}.

The resulting delay spectrum from a composite foreground model that combines diffuse and compact object emission consists of center--heavy features on all antenna spacings from compact objects and edge--heavy features from both components especially diffuse emission even on wide antenna spacings. Henceforth, we call this as the {\it pitchfork} structure in the {\it foreground} wedge.

All these observations are consistent with those from figures~\ref{fig:noiseless-dsm-delay-spectrum} and \ref{fig:noiseless-csm-delay-spectrum}. The observability of the predicted {\it pitchfork} signature depends on the relative levels of uncertainty in the foreground model and thermal noise. In our simulations, since thermal noise is $\gtrsim 10^7$~Jy~Hz, and features near the horizon delay limits are $\sim 10^7$~Jy~Hz, the {\it pitchfork} feature is not expected to be detected in a noisy scenario although this feature is marginally visible in the {\it zenith} pointing of the observed data (see figure~\ref{fig:fhd_data}). We attribute this to differences between our foreground model and the actual sky. However, with improvements to the foreground model and instrument configuration that can reduce thermal noise and other systematic uncertainties, we are confident the {\it pitchfork} signature in delay spectrum will be observed with high significance.

We also note that increasing the antenna spacing results in progressively improving the resolution along delay axis by increasing the number of delay bins inside the {\it foreground wedge}. This improves the localization of foreground objects whose signatures are imprinted in the delay spectrum. For instance, there is an increase in the number of secondary peaks in the delay spectrum between zero--delay and horizon delay limits as antenna spacing increases from $\sim$84~m to $\sim$170~m. In this case, these correspond to sidelobes of the power pattern that lie between the main lobe and the horizon along the North--South direction. In short antenna spacings, due to relatively lower resolution along delay axis inside the {\it foreground wedge} and a consequent loss of localization of direction of foreground emission, these secondary peaks blend in with other major peaks and are not distinctly visible. These findings are consistent with dicussion in \S\ref{sec:csm-side-lobe}.

In summary, the brightest signature ($\gtrsim 10^{10}$~Jy~Hz) is that of the Galactic center in the {\it off--zenith} pointing co--located at a westward sidelobe with a significantly high gain. The next brightest signature ($\sim 10^8$--$\sim 10^9$~Jy~Hz) is caused by compact emission appearing to be concentrated in the inner regions of the foreground wedge (near zero delays) rather than towards the edges. Diffuse emission co--located at the main lobe of the power pattern is fainter by approximately an order of magnitude relative to compact emission from the same region for a $\sim 84$~m antenna spacing but unlike the latter, diffuse emission decreases rapidly by over two orders of magnitude as antenna spacing is widened to $\sim 170$~m. However, diffuse emission is significantly high near horizon delay limits compared to main lobe of the power pattern as antenna spacing is widened. This leads to edge--heavy features in the delay spectrum. Complemented by the center-heavy compact foreground features, we have demosntrated the presence of a {\it three--pronged pitchfork}--shaped signature in the delay spectrum of foreground sky.

@@

The delay spectrum, $V_\eta(\overline{\mathbf{x}},\eta)$, not only carries information on spatial scales of foreground emission but also offers a unique advantage of viewing the sky through a combination of antenna spacing vector and the delay axis. This allows us to programmatically screen data for antenna spacings that are severly contaminated by foregrounds which can be weighted appropriately during data analysis. While the {\it foreground wedge} is a region of high foreground contamination, of particular interest to the EoR community is foreground contamination near horizon delay limits. These regions are responsible for significant spillover into the relatively clean {\it EoR window} due to spectral response of the instrument and foregrounds.

>>>>

Foregrounds leak into the EoR H{\sc i} power spectrum by either having intrinsic spectral variability, high instrumental chromaticity, or through approximations in power spectrum analysis.  Here we use existing catalogs and a high fidelity instrumental model to focus on the possible sources of instrumental chromaticity. 
To decouple from possible analysis effects, such as those pointed out by \citet{haz13}, we compute power spectra using the individual antenna pair based approach of \citet{pob13,par14}.  This approximates the power spectrum as the inverse fourier transform of the spectra generated by the correlator.

@@

We define the {\it delay spectrum} $V_\eta(\overline{\mathbf{u}},\eta)$ to be the inverse Fourier transform of $V_f(\overline{\mathbf{u}},f)$ along the frequency coordinate:
\begin{align}\label{eqn:delay-transform}
  V_\eta(\overline{\mathbf{u}},\eta) &= \int V_f(\overline{\mathbf{u}},f)\,W'_f(f)\,e^{i2\pi\eta f}\,\dif f,
\end{align}
where, $W'_f(f)$ is a frequency window weighting function which can be chosen to control the quality of the delay spectrum \citep{thy13,ved12}. $\eta$ represents the signal delay between the antenna pairs in an interferometer given by:
\begin{equation}\label{eqn:delay}
  \eta = \frac{\overline{\mathbf{x}}\cdot\overline{\mathbf{l}}}{c} = \frac{\overline{\mathbf{u}}\cdot\overline{\mathbf{l}}}{f}.
\end{equation}
Note that as the visibility is a fully complex quantity, its Fourier transform is fully Nyquist sampled over positive and negative delays.  Negative and positive delay\footnote{The terms delay and lag are used interchangeably in this paper.} correspond to the two hemispheres of the sky as transected by the antenna pair's baseline vector. For example a perfectly East-West baseline will see a source in the Eastern sky at positive delay, and conversely for a source in the Western sky.

@@

In order to make a quantitative comparison of delay spectra obtained with the MWA data and our simulations, we consider the following uncertainties. Our foreground models are derived from other higher frequency catalogs and sky maps. The inherent spread in spectral index increases the uncertainty while predicting fluxes at the observed frequency. Using simple error propagation, the fractional error in the delay spectrum caused by the spread in spectral index is $\sim \ln(f_\textrm{orig}/f)\,\Delta\alpha_\textrm{sp}$, where, $f_\textrm{orig}$ is the original frequency at which the catalog or map was created, $f=185$~MHz is the observed frequency, and $\Delta\alpha_\textrm{sp}$ is the spread (HWHM) in spectral index. From \citet{mau03}, we assume $\Delta\alpha_\textrm{sp} \approx 0.35$ for compact sources from NVSS and SUMSS catalogs. Although the model of \citet{deo08} yields a spectral index per direction on the sky, we assume similar uncertainties exist in spectral indices of our diffuse sky model as well, which is predominantly derived from 408~MHz map of \citet{has82}. Thus, fractional errors in delay spectrum from compact foreground objects and diffuse emission are $\sim$70\% and $\sim$30\% respectively. In addition, delay spectra from simulations and data each have uncertainties due to thermal noise with rms, $\Delta V^\textrm{N}_\eta(|\overline{\mathbf{x}}|,\eta) \sim$ 1.4$\times 10^7$~Jy~Hz, where the superscript N stands for thermal noise. We estimate the ratio of delay spectra from data and simulations as, $\rho = |V^\textrm{D}_\eta(|\overline{\mathbf{x}}|,\eta)|\,/\,|V^\textrm{S}_\eta(|\overline{\mathbf{x}}|,\eta)|$, where superscripts D and S denote data and simulation respectively. % Using the aforementioned uncertainties added in quadrature, we estimate the approximate expected uncertainty in $\log_{10}\rho$ and denote it by $\Delta\,\log_{10}\rho$.

@@

Figure~\ref{fig:data-sim-ratio} shows histograms of the distribution of $\log_{10}\rho$ for the {\it off--zenith} (top panel) and {\it zenith} (bottom panel) pointings respectively inside the {\it foreground wedge}. The median absolute deviation from these distributions correspond to $\sim$~90\% fractional difference between data and modeling on average in either case. This is in line with aforementioned uncertainties in foreground models and thermal noise in measurements. Currently we are significantly limited by unavailability of all--sky foreground models at the observed frequency of 185~MHz. Availability of an accurate foreground model will improve the agreement of simulations with observed data.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/delta_array_histogram_wedge_sim_data_log_ratio_0.3m_ground_custom_gaussian_FG_model_asm_all_sky_nside_64_Tsys_90.0K_185.0_MHz_30.7_MHz_bnw2.0.eps}
\caption{Histogram of logarithm of ratio of observed to simulated visibilities restricted to the {\it foreground wedge} --i.e. within the white lines of Figure \ref{fig:sim_data}-- for the {\it off--zenith} (top) and {\it zenith} (bottom) pointings. Median absolute deviations of $\log_{10}\rho$ are 0.29 and 0.27 respectively. A value of zero indicates when the ratio $\rho=1$. \label{fig:data-sim-ratio}}
% \includegraphics[width=\linewidth]{figures/v1_0/delta_array_histogram_wedge_sim_data_snr_log_ratio_0.3m_ground_custom_gaussian_FG_model_asm_all_sky_nside_64_Tsys_90.0K_185.0_MHz_30.7_MHz_bnw2.0.eps}
% \caption{Significance of logarithm of ratio of observed to simulated visibilities relative to expected uncertainty restricted to the {\it foreground wedge} for the {\it off--zenith} (top) and {\it zenith} (bottom) pointings. Probability densities and cumulative distributions are shown in black and gray respectively. Region inside the dotted vertical lines denotes when the ratio, $\rho$, is consistent with unity subject to expected uncertainty. Also indicated in percentages are confidence levels estimated conservatively for such a consistency. They are $\sim$~90\% and $\sim$~95\% for {\it off--zenith} and {\it zenith} pointings respectively. \label{fig:data-sim-ratio-snr}}
\end{figure}

@@

Having established that the model, by and large, matches the data, we proceed to examine in further detail the signatures seen in the modeled delay spectra.

@@

\item {\it Sky Model}: Our model of the sky consists of bright compact sources and diffuse emission on diverse spatial scales as seen from figures~\ref{fig:DSM} and \ref{fig:CSM}. It also consists of localized regions of strong emission such as the Galactic plane and Galactic center. The resulting sky emission is anisotropic. In fact, the patch of sky for MWA observations is chosen from regions of low foreground emission. In our detailed study, we have divided the transiting sky into four different ``bow-tie'' shaped regions with equal areas around zenith as the origin. This results in the ability to attribute the features in the delay spectrum to eight different sectors of the sky. %TODO possibly reference sections or figures here?

\item {\it Antenna Pair Orientation}: Since the spatial structure of our foreground model is not expected to be isotropic, we divide our antenna pairs by their orientation ($\theta_\textrm{b}$) measured counter-clockwise from East. We use the following bins: $-67\fdg 5\le \theta_\textrm{b} < -22\fdg 5$, $-22\fdg 5\le \theta_\textrm{b} < 22\fdg 5$, $22\fdg 5\le \theta_\textrm{b} < 67\fdg 5$, and $67\fdg 5\le \theta_\textrm{b} < 112\fdg 5$. The bin centers are oriented towards South-East, East, North-East, and North respectively. Since delays depend on antenna pair orientation, binning by orientation allows us to match delay spectrum features to different sky directions.

\item {\it Tile Pointing and Power Pattern}: Since the MWA tiles are steered electronically, the power pattern of the tile changes in any observing mode that tracks the source. The pointing system is capable of steering to points on a regular $\sim$7\arcdeg grid.  During observations we allow the sky to drift across the nearest available pointing, shifting between grid points as necessary. In our study, we take into account the effect of changing power pattern of the antennas on the delay spectrum (Figure~\ref{fig:power_pattern}). The LST of our data ranges from $\sim$~21 hours through $\sim$~2 hours, and for our analysis we chose {\it off--zenith} and {\it zenith} pointings shown in figure~\ref{fig:pointings}.

\item {\it Instrumental Passband}: The instrumental passband characteristics have a direct effect on the delay transform owing to the Fourier relationship between the two. Due to excision of noisy and aliased edge channels repeated in each coarse channel of our passband, we saw repeated patterns of the {\it foreground wedge} in the delay spectra. As already mentioned earlier, we have used the deconvolution algorithm ({\it CLEAN}) to rid the delay spectra of band shape effects. Since this paper focuses on studying the effects of foreground on the delay spectra, we do not pursue the characterization of artifacts from the band shape and imperfections in {\it CLEAN} deconvolution.

@@

%The foreground signatures we henceforth identify in the delay spectrum are characterized as arising out of the aforementioned factors. Due to numerous combinations of such factors, we use unique annotations to characterize these signatures.

It may be noted that numerous features overlap at varying levels of significance depending on the combinations of parameters. We assign the features to their predominant causes. Secondly, we have used noiseless cases to clearly illustrate the observed foreground signatures. With the addition of noise in the visibilities which is subject to observing time, some of the weaker features may not be as prominently visible. Since the foreground signatures are far too numerous and they reside in a vast volume of parameter space such as antenna pair orientation, power pattern, patch of sky under observation, and instrumental configuration, we highlight only some examples of the most notable features observed in the delay spectrum of foreground emission.

\subsection{Diffuse Foreground: Input model}\label{sec:diffuse}

@@

\subsubsection{Diffuse Emission Power Spectra}\label{sec:diffuse-features}

As we also see in Figure~\ref{fig:noiseless-dsm-delay-spectrum}, diffuse emission from within the field of view manifests in the {\it off--zenith} pointing as a branch at the top of the wedge, while in the {\it zenith} pointing it shows up at near zero delay. The former is seen at positive delays because the main lobe of the power pattern is centered eastward of zenith, whereas in the latter it is centered at zenith. As we see from Equation \ref{eqn:obsvis}, each baseline measures a single spatial mode on the sky with an angular size scale inverseley proportional to length of the baseline projected in the direction of emission.  Thus, on the {\it zenith} pointing, the signature of the smooth sky model, the horizontal line at zero delay, fades away on antenna spacings wider than $\gtrsim 125$~m because the sky model is devoid of spatial structures on scales $\lesssim$~0\fdg 75. On the off-zenith pointing we see power out to much longer baselines.

@@

Figure~\ref{fig:noiseless-csm-delay-spectrum} shows the delay spectrum compact source sky model for {\it off--zenith} and {\it zenith} pointings (shown in Figure~\ref{fig:CSM}) in the top and bottom panels respectively.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/delta_array_multi_baseline_CLEAN_noiseless_visibilities_0.3m_ground_custom_gaussian_FG_model_csm_all_sky_nside_64_Tsys_90.0K_185.0_MHz_30.7_MHz_bnw2.0.eps}
\caption{Delay spectrum amplitude, $|V_\eta(|\overline{\mathbf{x}}|,\eta)|$ (in units of Jy~Hz), obtained with simulations for {\it off--zenith} (top) and {\it zenith} (bottom) pointings for the compact objects foreground model without any thermal noise. White lines mark the boundaries of {\it foreground wedge} determined by the horizon delay limit and antenna spacing. The logarithmic color scale is identical to that in Figure~\ref{fig:noiseless-dsm-delay-spectrum}. Emission from compact objects gives rise to centrally concentrated features in the {\it foreground wedge}.\label{fig:noiseless-csm-delay-spectrum}}
\end{figure}

@@

\subsubsection{Compact Emission in Power Pattern Main Lobe}\label{sec:csm-main-lobe}
The response of an interferometer to a compact source is, to first order, flat with respect to baseline length. Therefor the distribution of sources within the primary field of view depends mainly on how the beam pattern maps to delay space for each antenna pair. Since the main lobe of the power pattern in the {\it off--zenith} pointing is centered eastward of zenith, the bulk of the compact foreground emission is seen in a branch with positive delay corresponding to the position of the pointing center. In the {\it zenith} pointing compact emission from the same patch of sky is seen as a bright horizontal arm at zero delay  since the main lobe of the power pattern is centered at zenith.

@@

\subsection{Combined Foreground Model}\label{sec:composite}

@@

When not dominated by the bright emission from the Galactic center, which happens only for certain pointings at certain times, the delay spectrum of the combined foreground model is dominated, in somewhat equal measure, by diffuse and compact emission. This is illustrated by a more detailed examination of the {\it zenith} pointing in our study.

Figure~\ref{fig:pitchfork-baselines} shows delay spectra of three antenna pairs of different antenna spacings oriented northward during the {\it zenith} pointing; each is a different vertical slice of the two dimensional power spectrum plots shown in Figures \ref{fig:fhd_data} et al. The diffuse, compact, and composite components are are shown as dotted, dashed, and solid lines respectively. The horizon delay limits are shown as vertical dot--dashed lines. The gray shaded area denotes the envelope of expected uncertainty in the delay spectrum. Sources inside the {\it foreground} wedge (between the horizon delay limits) are dominated by the $\sim$70\% error in predicting the spectral index of compact foreground objects, outside the wedge thermal noise dominates.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/delta_array_3_baseline_comparison_visibilities_0.3m_ground_custom_gaussian_FG_model_csm_all_sky_nside_64_Tsys_90.0K_185.0_MHz_30.7_MHz_bnw2.0.eps}
\caption{Absolute value of delay spectrum, $|V_\eta(\overline{\mathbf{x}},\eta)|$, for three chosen antenna spacings of length: $\sim$84~m ({\it top}), $\sim$105~m ({\it middle}), and $\sim$170~m ({\it bottom}). The antenna spacing vector is specified in each panel along East, North and Upward directions. The dotted, dashed and solid lines denote contributions from diffuse emission, compact objects and composite foreground model respectively. The dot--dashed lines mark the horizon delay limits. The shaded region denotes the envelope of uncertainty around expected values for composite model. This uncertainty is dictated by uncertainties in spectral indices of foreground models inside the delay horizon delay limits. Outside the horizon delay limits, it is predominantly due to thermal noise. Compact objects dominate the central regions of power spectrum while both components, especially diffuse emission on short antenna spacings, dominate near the horizon delay limits, giving rise to a {\it pitchfork}--shaped structure. \label{fig:pitchfork-baselines}}
\end{figure}

@@

Near the horizon delay limits at shorter antenna spacing, diffuse emission is brighter than that from compact objects.  Diffuse emission does not decrease as rapidly with increasing antenna spacing as was seen at zero--delay. In fact, even on widely spaced antennas, {\emph diffuse emission in the delay spectrum near the horizon delay limits exceeds that in the main lobe by more than two orders of magnitude}. This is due to apparent shortening of antenna spacings projected in the direction of smooth emission entering far from the primary field of view. This demonstrates the edge--heavy {\it two--pronged fork} feature discussed in \S\ref{sec:diffuse-long-baselines}.

The resulting baseline dependent power spectrum from a composite foreground model that combines diffuse and compact object emission consists of center--heavy features on all antenna spacings from compact objects and edge--heavy features from both components especially diffuse emission even on wide antenna spacings. Henceforth, we call this as the {\it pitchfork} structure in the {\it foreground} wedge.

All these observations are consistent with those from Figures~\ref{fig:noiseless-dsm-delay-spectrum} and \ref{fig:noiseless-csm-delay-spectrum}. The observability of the predicted {\it pitchfork} signature depends on the relative levels of uncertainty in the foreground model and thermal noise. In our simulations, since thermal noise in these very short snapshots is $\gtrsim 10^7$~Jy~Hz, and features near the horizon delay limits are $\sim 10^7$~Jy~Hz, the {\it pitchfork} feature is not expected to be detected in a noisy scenario although this feature is marginally visible in the {\it zenith} pointing of the observed data (see Figure~\ref{fig:fhd_data}). We attribute this to differences between our foreground model and the actual sky. However, with improvements to the foreground model and the addition of more data to reduce the thermal noise, we are confident the {\it pitchfork} signature in delay spectrum will be observed with high significance.

We also note that increasing the antenna spacing results in progressively improving the resolution along delay axis by increasing the number of delay bins inside the {\it foreground wedge}. This improves the localization of foreground objects whose signatures are imprinted in the delay spectrum. For instance, there is an increase in the number of secondary peaks in the delay spectrum between zero--delay and horizon delay limits as antenna spacing increases from $\sim$84~m to $\sim$170~m. In this case, these correspond to sidelobes of the power pattern that lie between the main lobe and the horizon along the North--South direction. In short antenna spacings, due to relatively lower resolution along delay axis inside the {\it foreground wedge} and a consequent loss of localization of direction of foreground emission, these secondary peaks blend in with other major peaks and are not distinctly visible. These findings are consistent with the discussion in \S\ref{sec:csm-side-lobe}.

In summary, the brightest signature ($\gtrsim 10^{10}$~Jy~Hz) is that of the Galactic center in the {\it off--zenith} pointing co--located at a westward sidelobe with a significantly high gain. The next brightest signature ($\sim 10^8$--$\sim 10^9$~Jy~Hz) is caused by compact emission appearing to be concentrated in the inner regions of the foreground wedge (near zero delays) rather than towards the edges. Diffuse emission co--located at the main lobe of the power pattern is fainter by approximately an order of magnitude relative to compact emission from the same region for a $\sim 84$~m antenna spacing but unlike the latter, diffuse emission decreases rapidly by over two orders of magnitude as antenna spacing is widened to $\sim 170$~m. However, diffuse emission is significantly high near horizon delay limits compared to main lobe of the power pattern as antenna spacing is widened. This leads to edge--heavy features in the delay spectrum. Complemented by the center-heavy compact foreground features, we have demonstrated the presence of a {\it three--pronged pitchfork}--shaped signature in the delay spectrum of foreground sky.

@@

The delay spectrum, $V_\eta(\overline{\mathbf{x}},\eta)$, not only carries information on spatial scales of foreground emission but also offers a unique advantage of viewing the sky through a combination of antenna spacing vector and the delay axis. This allows us to programmatically screen data for antenna spacings that are severely contaminated by foregrounds which can be weighted appropriately during data analysis. While the {\it foreground wedge} is a region of high foreground contamination, of particular interest to the EoR community is foreground contamination near horizon delay limits. These regions are responsible for significant spillover into the relatively clean {\it EoR window} due to spectral response of the instrument and foregrounds.
